---
title: "Rhinovirus subtypes with NPA microbiome in MARC-35"
date: "`r Sys.Date()`"
output: html_document
---
# Study design
  **Samples**: MARC-35, acute/chronic cohort.
  **Exposure**: RV subtypes (A, B, C) vs. solo RSV (reference).
  **Outcome**: NPA microbiota (PAM clusters, top 10 genera etc)
  **Analysis**: 
  **Sensitivity analysis**: e.g., remove RSV-RV coinfection (but it'll lose the power)
  **References**: Bashir et al. JACI 2018 (https://www.ncbi.nlm.nih.gov/pubmed/29111214)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
version
```

## Analysis
## 1-1. Uploading the packages
```{r Upload packages, message=FALSE}
library("tidyverse")
library("stringr")  # Work on string
library("purrr")
library("tableone")
library("lme4")     # for GLM
```

## 1-2.  Importing m35 metadata
```{r Import metadata, message=FALSE}
library("readstata13")
df_m35 <- read.dta13("../../severity/data/m35/m35_metadata_n1016_2018march.dta")
#colnames(df_m35)
dim(df_m35) 
# Limit variables
metadata_pam <- df_m35 %>% 
                  select(-c(acute_cohort, enrolled_subj, site_region:site_region_lat,
                            collection_season_NPA, breastmilk_mostly_6mo:mo12_solidfood,
                            intake_overnight_hosp, intake_antibiotics_length,
                            index_bp_woari, mata_ppi, intake_adultsinhome, mo30_cat_0to1:mo30_farmanimal_0to1,
                            intch_retractions, intch_oralintake:intch_hypertonic_saline,
                            inpatient_hfo:inpatient_hel, SampleID_NPA,
                            Adenovirus:hMPV, myco_pneumoniae:RSVB, pathogen_sum,
                            RnaseP:RSVB_CT, hMPV_CT:myco_pneumoniae_CT,
                            alb_gdl,
                            HeirarchicalWard4:HeirarchicalAvg5)) %>%  # remove unnecessary variables
                  arrange(., by=study_id)
# Rename HRV --> RV
metadata_pam$RV  <- as.factor(metadata_pam$HRV)   
  table(metadata_pam$RV, useNA="ifany"); class(metadata_pam$HRV)  # 214 RV by PCR
```


## 1-3. Recoding the metadata variables of m35
```{r Re-code metadata of m35, message=FALSE}
# Covariates    
    # Age
        metadata_pam$Agecat2 <-NA
        metadata_pam$Agecat2[metadata_pam$Age_mo>=0 & metadata_pam$Age_mo <6] <- 1
        metadata_pam$Agecat2[metadata_pam$Age_mo>=6] <- 2
            metadata_pam$Agecat2 <- as.factor(metadata_pam$Agecat2)
            table(metadata_pam$Agecat2, useNA="ifany"); class(metadata_pam$Agecat2)   
        metadata_pam$Agecat3 <-NA
        metadata_pam$Agecat3[metadata_pam$Age_mo>=0 & metadata_pam$Age_mo <2] <- 1
        metadata_pam$Agecat3[metadata_pam$Age_mo>=2 & metadata_pam$Age_mo <6] <- 2
        metadata_pam$Agecat3[metadata_pam$Age_mo>=6] <- 3
            metadata_pam$Agecat3 <- as.factor(metadata_pam$Agecat3)
            table(metadata_pam$Agecat3, useNA="ifany"); class(metadata_pam$Agecat3)           
    # Sex  
        metadata_pam$female = metadata_pam$intake_sex
            table(metadata_pam$female, useNA="ifany"); class(metadata_pam$female)  
                
    # Race/ethnicity
        metadata_pam$raceethn <- as.factor(metadata_pam$raceethn)
            table(metadata_pam$raceethn, useNA="ifany"); class(metadata_pam$raceethn)  
    # Parental asthma (missing n=1)
        metadata_pam$parent_asthma = factor(metadata_pam$parent_asthma)
            table(metadata_pam$parent_asthma, useNA="ifany"); class(metadata_pam$parent_asthma)  
    # Maternal smoke  (missing n=1)
        metadata_pam$mata_smoke = factor(metadata_pam$mata_smoke)
            table(metadata_pam$mata_smoke, useNA="ifany"); class(metadata_pam$mata_smoke)          
    # Fall birth (Sep, Oct, Nov) -- https://www.ncbi.nlm.nih.gov/pubmed/18776151
        metadata_pam$birth_season <- factor(metadata_pam$birth_season)
            table(metadata_pam$birth_season, useNA="ifany"); class(metadata_pam$birth_season) 
        metadata_pam$fall_birth <- ifelse(metadata_pam$birth_season==3, TRUE, FALSE)
            table(metadata_pam$fall_birth, useNA="ifany"); class(metadata_pam$fall_birth)
    # Prematurity
        metadata_pam$premature37 <- factor(metadata_pam$premature37)
            table(metadata_pam$premature37, useNA="ifany"); class(metadata_pam$premature37)
    # Birth mode (1 missing)
        metadata_pam$mata_delivery <- factor(metadata_pam$mata_delivery)
            table(metadata_pam$mata_delivery, useNA="ifany"); class(metadata_pam$mata_delivery)
    # low birth wt  < 5 lb
        metadata_pam$intake_child_weight_born <- as.numeric(metadata_pam$intake_child_weight_born)
            table(metadata_pam$intake_child_weight_born, useNA="ifany");    class(metadata_pam$intake_child_weight_born)    
        metadata_pam$low_birth_wt <-NA
        metadata_pam$low_birth_wt[metadata_pam$intake_child_weight_born==1] <-TRUE
        metadata_pam$low_birth_wt[metadata_pam$intake_child_weight_born==2] <-TRUE
        metadata_pam$low_birth_wt[metadata_pam$intake_child_weight_born==3] <-FALSE
        metadata_pam$low_birth_wt[metadata_pam$intake_child_weight_born==4] <-FALSE    
            table(metadata_pam$low_birth_wt, useNA="ifany"); class(metadata_pam$low_birth_wt)
    # Previous breathing
        metadata_pam$prev_breathingprob2 <-NA
        metadata_pam$prev_breathingprob2[metadata_pam$prev_breathingprob==0] <-FALSE
        metadata_pam$prev_breathingprob2[metadata_pam$prev_breathingprob==1] <-TRUE
        metadata_pam$prev_breathingprob2[metadata_pam$prev_breathingprob==2] <-TRUE
            table(metadata_pam$prev_breathingprob2, useNA="ifany"); class(metadata_pam$prev_breathingprob2) 
    # H/o eczema 
        metadata_pam$intake_eczema_history <- factor(metadata_pam$intake_eczema_history)
            table(metadata_pam$intake_eczema_history, useNA="ifany"); class(metadata_pam$intake_eczema_history)   
    #Post-natal smoking exposure
        metadata_pam$intake_smoke <- factor(metadata_pam$intake_smoke)    
            table(metadata_pam$intake_smoke, useNA="ifany"); class(metadata_pam$intake_smoke) 
    # Siblings
        metadata_pam$children_home2 <-NA
        metadata_pam$children_home2[metadata_pam$numb_children_home==0] <-FALSE
        metadata_pam$children_home2[metadata_pam$numb_children_home!=0] <-TRUE
            table(metadata_pam$children_home2, useNA="ifany"); class(metadata_pam$children_home2)           
            
    # breastmilk_mostly_3mo (10 missing)
        metadata_pam$breastmilk_mostly_3mo <- factor(metadata_pam$breastmilk_mostly_3mo)    
            table(metadata_pam$breastmilk_mostly_3mo, useNA="ifany"); class(metadata_pam$breastmilk_mostly_3mo)            
            
        metadata_pam$breastmilk_mostly_3mo9 <-NA
        metadata_pam$breastmilk_mostly_3mo9[metadata_pam$breastmilk_mostly_3mo==0] <-0
        metadata_pam$breastmilk_mostly_3mo9[metadata_pam$breastmilk_mostly_3mo==1] <-1
        metadata_pam$breastmilk_mostly_3mo9[is.na(metadata_pam$breastmilk_mostly_3mo)]<-9        
            metadata_pam$breastmilk_mostly_3mo9 <- as.factor(metadata_pam$breastmilk_mostly_3mo9)
            table(metadata_pam$breastmilk_mostly_3mo9, useNA="ifany"); class(metadata_pam$breastmilk_mostly_3mo9) 
          
    # Corticosteroid use before the index hospitalization
        metadata_pam$corticosteroids_life <- as.factor(metadata_pam$corticosteroids_life)   
            table(metadata_pam$corticosteroids_life, useNA="ifany"); class(metadata_pam$corticosteroids_life)          
    # Corticosteroid use during the pre-hospitalization visit
        metadata_pam$corticosteroids_pre <- as.factor(metadata_pam$corticosteroids_pre)   
            table(metadata_pam$corticosteroids_pre, useNA="ifany"); class(metadata_pam$corticosteroids_pre)         
            
    # Wt at presentation
                    summary(metadata_pam$intch_weight)
        metadata_pam$intch_weight_cat2 = 
                    ifelse(metadata_pam$intch_weight<5.95,  "Low",
                    ifelse(metadata_pam$intch_weight>=5.95, "High",
                    NA))
            metadata_pam$intch_weight_cat2 <- as.factor(metadata_pam$intch_weight_cat2)
            table(metadata_pam$intch_weight_cat2, useNA="ifany"); class(metadata_pam$intch_weight_cat2)
          
    # sIgE (food or aero)
        metadata_pam$IgE_any_food <- as.factor(metadata_pam$IgE_any_food)   
            table(metadata_pam$IgE_any_food, useNA="ifany"); class(metadata_pam$IgE_any_food)  
        metadata_pam$aeroallergens_atopy <- as.factor(metadata_pam$aeroallergens_atopy)   
            table(metadata_pam$aeroallergens_atopy, useNA="ifany"); class(metadata_pam$aeroallergens_atopy)  
        metadata_pam$IgE_cat2 <- ifelse(metadata_pam$IgE_any_food==1 | metadata_pam$aeroallergens_atopy==1, TRUE,
                                    FALSE)
            table(metadata_pam$IgE_cat2, useNA="ifany"); class(metadata_pam$IgE_cat2) 
     
    # O2 sat (3 missing)
        metadata_pam$o2sat_initial <- as.factor(metadata_pam$o2sat_initial)   
            table(metadata_pam$o2sat_initial, useNA="ifany"); class(metadata_pam$o2sat_initial)  
    # Severity
        metadata_pam$CPAPintubate <- as.factor(metadata_pam$CPAPintubate)   
            table(metadata_pam$CPAPintubate, useNA="ifany"); class(metadata_pam$CPAPintubate)   
        metadata_pam$inpatient_location___3 <- as.factor(metadata_pam$inpatient_location___3)   
            table(metadata_pam$inpatient_location___3, useNA="ifany"); class(metadata_pam$inpatient_location___3)   
        metadata_pam$IntensiveTreatment <- as.factor(metadata_pam$IntensiveTreatment)   
            table(metadata_pam$IntensiveTreatment, useNA="ifany"); class(metadata_pam$IntensiveTreatment) 
        metadata_pam$LOS_dys <- as.integer(metadata_pam$LOS_dys)   
            summary(metadata_pam$LOS_dys, useNA="ifany"); class(metadata_pam$LOS_dys)      
        metadata_pam$LOS3dys <- as.factor(metadata_pam$LOS3dys)   
            table(metadata_pam$LOS3dys, useNA="ifany"); class(metadata_pam$LOS3dys)   
        
    # eos_4perc  (16 missing)
        metadata_pam$eos_4perc <- as.factor(metadata_pam$eos_4perc)   
            table(metadata_pam$eos_4perc, useNA="ifany"); class(metadata_pam$eos_4perc)          
    # Microbiota clusters (PAM 4)
        metadata_pam$PAM4 = 
                    ifelse(metadata_pam$PAM4==1, "Moraxella dominant",
                    ifelse(metadata_pam$PAM4==2, "Haemophilus dominant",
                    ifelse(metadata_pam$PAM4==3, "Mixed",
                    ifelse(metadata_pam$PAM4==4, "Streptococcus dominant",
                    NA))))
          table(metadata_pam$PAM4, useNA="ifany")         
    # site
        metadata_pam$site <- factor(metadata_pam$site)
            table(metadata_pam$site, useNA="ifany"); class(metadata_pam$site)  
    # Rec wheezing
          metadata_pam$recurrent_wheeze_sleep_36mo_cat <- factor(metadata_pam$recurrent_wheeze_sleep_36mo)
            table(metadata_pam$recurrent_wheeze_sleep_36mo_cat, useNA="ifany"); class(metadata_pam$recurrent_wheeze_sleep_36mo_cat)                    
          #metadata_pam$recurrent_wheeze_36mo_cat <- factor(metadata_pam$recurrent_wheeze_36mo)
          #  table(metadata_pam$recurrent_wheeze_36mo_cat, useNA="ifany"); class(metadata_pam$recurrent_wheeze_36mo_cat)
          #metadata_pam$ICSuse_36mo_cat <- factor(metadata_pam$ICSuse_36mo)
          #  table(metadata_pam$ICSuse_36mo_cat, useNA="ifany"); class(metadata_pam$ICSuse_36mo_cat)   

            
    # Virology        
        metadata_pam$RSVonly <- as.factor(metadata_pam$RSVonly)   
            table(metadata_pam$RSVonly, useNA="ifany"); class(metadata_pam$RSVonly)   
        metadata_pam$RVonly<- as.factor(metadata_pam$RVonly)   
            table(metadata_pam$RVonly, useNA="ifany"); class(metadata_pam$RVonly)         

        #metadata_pam$RSVA <- as.factor(metadata_pam$RSVA)   
        #    table(metadata_pam$RSVA, useNA="ifany"); class(metadata_pam$RSVA)  
        #metadata_pam$RSVB <- as.factor(metadata_pam$RSVB)   
        #    table(metadata_pam$RSVB, useNA="ifany"); class(metadata_pam$RSVB)
        metadata_pam$RSV <- as.factor(metadata_pam$RSV)   
            table(metadata_pam$RSV, useNA="ifany"); class(metadata_pam$RSV) 
            
        metadata_pam$RV  <- as.factor(metadata_pam$HRV)   
            table(metadata_pam$RV, useNA="ifany"); class(metadata_pam$HRV)  
        metadata_pam$OTHERvirus <-NA
            metadata_pam$OTHERvirus[metadata_pam$RSV==1 | metadata_pam$RV==1] <-FALSE    
            metadata_pam$OTHERvirus[metadata_pam$RSV !=1 & metadata_pam$RV !=1] <-TRUE
            table(metadata_pam$OTHERvirus, useNA="ifany"); class(metadata_pam$OTHERvirus)             
```

## 1-3 RV count in m35
```{r RV count in m35}
# RV subtypes
  table(metadata_pam$RVA_NPAseq, useNA="ifany") # n=97
  table(metadata_pam$RVB_NPAseq, useNA="ifany") # n=15
  table(metadata_pam$RVC_NPAseq, useNA="ifany") # n=99
  table(metadata_pam$RVA_NPAseq, metadata_pam$RVB_NPAseq, useNA="ifany") # A+B, n=1
  table(metadata_pam$RVA_NPAseq, metadata_pam$RVC_NPAseq, useNA="ifany") # A+C, n=4
  table(metadata_pam$RVB_NPAseq, metadata_pam$RVC_NPAseq, useNA="ifany") # B+C, n=2

# Create RV sutbype variable
  metadata_pam$RVsubtype <- "Neither ABC"
  metadata_pam$RVsubtype[metadata_pam$RVA_NPAseq==1 & metadata_pam$RVB_NPAseq==0 & metadata_pam$RVC_NPAseq==0] <-"Sole RVA"
  metadata_pam$RVsubtype[metadata_pam$RVA_NPAseq==0 & metadata_pam$RVB_NPAseq==1 & metadata_pam$RVC_NPAseq==0] <-"Sole RVB"
  metadata_pam$RVsubtype[metadata_pam$RVA_NPAseq==0 & metadata_pam$RVB_NPAseq==0 & metadata_pam$RVC_NPAseq==1] <-"Sole RVC"
  metadata_pam$RVsubtype[metadata_pam$RVA_NPAseq==1 & metadata_pam$RVB_NPAseq==1 & metadata_pam$RVC_NPAseq==0] <-"Coinfection"
  metadata_pam$RVsubtype[metadata_pam$RVA_NPAseq==0 & metadata_pam$RVB_NPAseq==1 & metadata_pam$RVC_NPAseq==1] <-"Coinfection"
  metadata_pam$RVsubtype[metadata_pam$RVA_NPAseq==1 & metadata_pam$RVB_NPAseq==0 & metadata_pam$RVC_NPAseq==1] <-"Coinfection"
  metadata_pam$RVsubtype[metadata_pam$RVA_NPAseq==1 & metadata_pam$RVB_NPAseq==1 & metadata_pam$RVC_NPAseq==1] <-"Coinfection"
    metadata_pam$RVsubtype <- as.factor(metadata_pam$RVsubtype)
    table(metadata_pam$RVsubtype , useNA="ifany")
```

## 1-4 Create subpopultions for m35
```{r Create subpopulations for m35}
# Limit to RVA, RVB, RVC, and RSV-only, with PAM (microbiota data) -- multiple RV species (n=7 excluded)
  df_m35_ABCRSV <- metadata_pam %>% 
                    filter(RVsubtype=="Sole RVA" | RVsubtype=="Sole RVB" | RVsubtype=="Sole RVC" | RSVonly==1) %>%
                    filter(!is.na(PAM4))
  table(df_m35_ABCRSV$RVsubtype, useNA="ifany")
  nrow(df_m35_ABCRSV)

# New variables for the 4 virus categories
  df_m35_ABCRSV$virus_cat4 <- "RSV-only"
  df_m35_ABCRSV$virus_cat4[df_m35_ABCRSV$RVsubtype=="Sole RVA"] <-"Sole RVA"
  df_m35_ABCRSV$virus_cat4[df_m35_ABCRSV$RVsubtype=="Sole RVB"] <-"Sole RVB"
  df_m35_ABCRSV$virus_cat4[df_m35_ABCRSV$RVsubtype=="Sole RVC"] <-"Sole RVC"
    df_m35_ABCRSV$virus_cat4 <- as.factor(df_m35_ABCRSV$virus_cat4)
    table(df_m35_ABCRSV$virus_cat4, useNA="ifany")
```

## 2-1. Descriptive (overall)
```{r Overall characteristics of m35}
summary(df_m35_ABCRSV$Age_mo)
table(df_m35_ABCRSV$female)
table(df_m35_ABCRSV$raceethn)
table(df_m35_ABCRSV$RSV, df_m35_ABCRSV$virus_cat4)
table(df_m35_ABCRSV$virus_cat4, df_m35_ABCRSV$RSV, useNA="ifany") # MANY RSV/RV CO-INFECTION
```
    
## 2-2. Table 1 for m35
```{r Table 1 of m35, eval=TRUE, message=FALSE}
    #colnames(df_m35_ABCRSV)
    table1_tbl <- df_m35_ABCRSV %>% select (-c(chronic_cohort, intake_sex,  intake_insurance, intake_income, inpatient_intub, inpatient_cpap,
                                            bioD, freeD, dbp_ugml, HRV, RVonly, NoPathogens_NPA, RVA_NPAseq:RVC_NPAseq, genotype_RV_NPAseq,
                                            milk_ISAC:total_conc_ku_l_r))  # For Table 1, remove some variables
    vars <- names(table1_tbl)[-1]
    table_1<-CreateTableOne(vars = vars, strata = c("virus_cat4"), 
                           data=table1_tbl,
                           factorVars = c("female", "raceethn", 
                                         "parent_asthma", "parent_eczema" , "mata_smoke",
                                         "admit_month", "admit_year",
                                         "premature37","intake_child_weight_born", "mata_delivery",
                                         "prev_breathingprob2", "intake_eczema_history",
                                         "intake_smoke", "intake_daycare", "children_home2","breastmilk_mostly_3mo",
                                         "corticosteroids_life", "corticosteroids_pre",
                                         "IgE_cat2", "IgE_any_food", "aeroallergens_atopy","eos_4perc",
                                         "RSVonly", "RVonly", "OTHERvirus",
                                         "RSV", "RV", "RVsubtype",
                                          "PAM4", 
                                         "o2sat_initial", 
                                         "CPAPintubate", "inpatient_location___3","IntensiveTreatment", "LOS3dys",
                                          "recurrent_wheeze_sleep_36mo"
                                         ),
                           includeNA = TRUE)  # if TRUE, NA as a category (ie, no exclusion)
          print(table_1, nonnormal = c("Age_mo", "intch_weight","LOS_dys",
                                      "intch_respiratory_rate", "Shannon", "Simpson", "ObservedOTUs",
                                      "RANTES_pgml","LL37_ngml", "vitd25_ngml",  "LL37_ngml",
                                      "HRV_CT",
                                      "RWsleep_time2event_36mo" ),
                            exact = c("intake_icu_born", "mata_antibiotics_labor", "mata_antibiotics_b4labor",
                                      "intake_rsvshot"), 
                            smd = FALSE,
                            quote=TRUE, noSpaces = TRUE,
                            max.print=TRUE)         
```                     
 
 
 ## For genus-level analysis (e.g., box-plot) by RV subtype, let's consider to use **violin plot**: 
 https://www.r-graph-gallery.com/violin-plot/
 
 